---
title: "Problem Set 2"
subtitle: "Due date: 18 September"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
  warning: false
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 25**

## Question 1

*Points: 10*

::: callout-note
5/10
:::

Suppose I want to examine the hypothesis that consuming political news coverage from cable network channels (e.g., CNN, MSNBC, Fox News, etc.) causes citizens to have less trust in government than if they were to consume it from mainstream print media (e.g., New York Times, Wall Street Journal, etc.). What is more, suppose that I believe that this effect is greatest among "young" citizens. Design the contours of an experiment to test this proposition (and be sure to sketch the key elements of the experimental design). Second, briefly describe any significant shortcomings that the experimental design might have.

::: callout-note
This is a well thought out answer; however, you are missing some important elements of an experiment.

1.  You needed to control the distribution of your treatment (getting political news from print media or from cable networks). At the moment, your participants decide this.

2.  Therefore, you only have a control group. You needed to randomly assign your participants to one of three groups: control, print media treated, cable network treated. Get those in the treatment groups to restrict their sources of political news to their assigned source.
:::

First, we need to draw a sample from a larger population. Then begin with asking respondents if they consume political news coverage from cable network or print media - here, we should inform the respondents that they can only choose one option. So we have a sub-sample of "cable network" and another one of "print media". We continue by asking respondents to identify their age. For this experiment, it may be better if we sort respondents into age groups, i.e., "below 35", "35-55", and "above 55", so that it gives us a clearer characterization of "young" citizens. And we move on to asking respondents about their level of trust in government with answers like "6-Mostly trust", "5-Generally trust", "4-Somewhat trust", "3-Somewhat don't trust", "2-Generally don't trust", and "1-Barely trust".

Now we can calculate two means for the level of trust, one for "cable network", the other for "print media". We can compare the difference between the two means and see if it supports our first hypothesis. We can look further by calculating the means of respective age groups from both sub-samples and see if the difference in means is the largest for young citizens, which in turn help us examine the second hypothesis.

Potential shortcomings:

<<<<<<< HEAD
1\) It may be the case that citizens consume political news coverage from multiple/mixed sources, though they would have a preference. And thus, we only know the factual outcome by assigning them into sub-samples. It may be a necessary step to take before we move forward to actually start the experiment, but there's also trade-off by making assignment this way.

2\) The level of trust may depend on a variety of factors, and there may also be confounder(s) affecting both, including the proximity of residence - suburban areas may find it difficult to access cable network and thus have to depend on print network. In such case, people who don't have cable services are systematically different from those have. Also, media have their own political stance where they have selection biases on which materials to be covered in their news coverage.

3\) This may be irrelevant but it seems to me that the source of information does not affect people's attitude toward government as the way we would expect. Real world surveys have shown how Democrats/Democratic-leaning voters are in favour of Democratic politicians regardless how they are doing in office, and the scenario is quite similar for Republican/Republican-leaning voters. This could undermine the credibility of our assumptions in the first place.
=======
1)  It may be the case that citizens consume political news coverage from multiple/mixed sources, though they would have a preference. And thus, we only know the factual outcome by assigning them into sub-samples. It may be a necessary step to take before we move forward to actually start the experiment, but there's also trade-off by making assignment this way.

2)  The level of trust may depend on a variety of factors, and there may also be confounder(s) affecting both, including the proximity of residence - suburban areas may find it difficult to access cable network and thus have to depend on print network. In such case, people who don't have cable services are systematically different from those have. Also, media have their own political stance where they have selection biases on which materials to be covered in their news coverage.

3)  This may be irrelevant but it seems to me that the source of information does not affect people's attitude toward government as the way we would expect. Real world surveys have shown how Democrats/Democratic-leaning voters are in favour of Democratic politicians regardless how they are doing in office, and the scenario is quite similar for Republican/Republican-leaning voters. This could undermine the credibility of our assumptions in the first place.

$$
Having control and treatment groups is CORE to this, otherwise it's NOT an experiment.

Control group: individuals who continue to consume political news with their own source of choice
Treatment group: 1) Cable Network only, and 2) Print Media only
Then ask participants about their level of trust in the government. Could run for a pretty long period of time (e.g., one year).

Shortcomings: 1) systematically different from population; 2) group may not be identical, in particular, the age; 3) compliance - source of news
$$
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

## Question 2

*Points: 10*

::: callout-note
4.5/10
:::

$$
Sampling variability: a statistic refers to how much the statistics varies from samples to samples
$$

Below are four hypothetical research designs to determine the effects of democratic governance on the funding of educational institutions. Briefly evaluate each sampling design in terms of its: (1) sampling variability; and, (2) sources of potential bias.

A.  A sample of educational funding levels for every country in Europe for each of the last 100 years.

<<<<<<< HEAD
    The sampling variability tends to be huge, given that we are looking at every European country over a time period of 100 years.
=======
::: callout-note
**Sampling variability:** incorrect. The sampling variability will be small because we are drawing every country from our limited pool of countries.

$$
For this scenario, we actually have ZERO sample variablity as all samples are identical.
Bias: towards the relationship present in European countries
$$

**Sampling bias:** partially correct. There is no random sampling error (because there is no random selection).
:::

The sampling variability tends to be huge, given that we are looking at every European country over a time period of 100 years.
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

    Potential bias may include: 1) Over the last 100 years, a number of historical events occurred in Europe, particularly the World War and the Cold War. They both tend to be confounding variables that influence both democratic governance and education funding level. 2) Random sampling error, in which a country with a small population and a small GDP may spend more funding on educational institutions in terms of percentage points of its GDP.

B.  A sample of educational funding levels in five randomly picked countries drawn from the global population of countries for each of the last 10 years.

<<<<<<< HEAD
    The sampling variability may be large or small, depending on the random picks of countries. If all or most of these randomly picked countries are well-developed democratic countries with good governance, the variability could be small. However, the likelihood of getting huge variability in this scenario may be much higher than geting a small variability.
=======
::: callout-note
**Sampling variability:** incorrect. Sampling variability will be large: we are selecting such a small proportion of our population.

**Sampling bias:** correct.
:::

$$
No bias, because every country has equal chance to be selected.
$$

The sampling variability may be large or small, depending on the random picks of countries. If all or most of these randomly picked countries are well-developed democratic countries with good governance, the variability could be small. However, the likelihood of getting huge variability in this scenario may be much higher than geting a small variability.
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

    Potential bias may include: Random sampling error, in which the sample statistics differ by chance from the population parameter. This leads to possible huge or small variability depending on the countries picked.

C.  A survey of educational administrators that yields responses that allow for a comparison of educational funding levels in Russia and the United States for the six-year period from 2010 to 2015.

<<<<<<< HEAD
    The sampling variability may be huge as geographical proximity of an education institution would have a large effect on administrators - city vs. surburbs, wealthy states/provinces vs. low-income states/provinces. In the United States, there is also variability between public institutions and private institutions where the latter tend to be more financially persistant.
=======
::: callout-note
**Sampling variability:** incorrect. It would be big if they only managed to survey a small proportion of administrators.

**Sampling bias:** partially correct. We are interested in the effects of democratic governance on the funding of educational institutions globally. Restricting our sample to two countries biases us away from seeing this relationship across all democracies vs. non-democracies.
:::

$$
Sampling variablity: Ideally, it should be small because every time it's the exactly same two groups of people. But if only a small proportion of education administrators repond to the survey, sampling variability could be huge.
Bias: towards the relationship of the US and Russia
$$

The sampling variability may be huge as geographical proximity of an education institution would have a large effect on administrators - city vs. surburbs, wealthy states/provinces vs. low-income states/provinces. In the United States, there is also variability between public institutions and private institutions where the latter tend to be more financially persistant.
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

    Potential bias may include: 1) Selection bias where educational administrators in different regions may not have an equal chance to be surveyed. 2) Response bias where some educational administrators are more likely to be measured.

D.  A sample of 75 randomly selected countries' educational funding levels for each of the past 25 years.

<<<<<<< HEAD
    The sampling variability may again be either huge or small depending on the countries selected, but the extent may not be that significant in Scenario B since we have a larger sample (75 to 5) and a longer measuring time period.
=======
::: callout-note
**Sampling variability:** partially correct. Sampling variability does not depend on a single draw of countries.

**Sampling bias:** correct.
:::

$$
Sampling viriability: should be small, 75 countries roughly account for 40% of all countries.
Bias: no bias as each country has equal chance to be selected.
$$

The sampling variability may again be either huge or small depending on the countries selected, but the extent may not be that significant in Scenario B since we have a larger sample (75 to 5) and a longer measuring time period.
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

    Potential bias may include: Random sampling error where we still have the chances of running into sample statistics differing by chance, though the extent may be smaller with a larger sample size.

## Question 3

*Points: 5*

The COVID-19 pandemic wreaked havoc on the global economy. People became sick or died, supply chains snapped, and we all sheltered behind strict national borders or in our homes. We can see the effect of these changes when we look at how the growth of countries' Gross Domestic Product (GDP) changed between 2019 and the global pandemic's start in 2020.

::: callout-note
GDP is a popular measure of a country's annual economic output. Its annual growth tends to be approximately Normally distributed.
:::

```{r}
library(tidyverse)
library(wbstats)

gdp_growth_df <- wb_data("NY.GDP.MKTP.KD.ZG", 
                         start_date = 2019, 
                         end_date = 2020, 
                         return_wide = F)

ggplot(gdp_growth_df, aes(x = value, fill = factor(date))) + 
  geom_density(alpha = 0.75) + 
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold"),
        plot.title.position = "plot") + 
  labs(title = "Countries' annual percentage growth rate of GDP in 2019 and 2020",
       x = "Annual growth in GDP (%)",
       y = "Density",
       fill = "Year",
       caption = "Source: World Bank")
```

### Part A

*Points: 2*

::: callout-note
2/2
:::

Describe the distribution of all countries' annual GDP growth rates in 2019 (found in the `gdp_growth_df` dataframe). Do the same for 2020. Include in those descriptions their five-number summaries.

```{r}
library(tidyverse)
library(skimr)

gdp2019 <- gdp_growth_df %>%
  select(country, date, value) |>
  filter(date %in% c("2019"))
skim(gdp2019, value)

gdp2020 <- gdp_growth_df %>%
  select(country, date, value) |>
  filter(date %in% c("2020"))
skim(gdp2020, value)
```

<<<<<<< HEAD
Both GDP growth rates of 2019 and 2020 are roughly normally distributed, while year 2019 was more clustered and year 2020 was more scattered. The five-number summary of year 2019 is as follows: min = -11.32, Q1 = 1.14, median = 2.70, Q3 = 4.76, max = 23.41. The five-number summary of year 2020 is as follows: min = -58.32, Q1 = -8.28, median = -3.87, Q3 = -1.10, max = 43.48
=======
Both GDP growth rates of 2019 and 2020 are roughly normally distributed. The five-number summary of year 2019 is as follows: min = -11.32, Q1 = 1.14, median = 2.70, Q3 = 4.76, max = 23.41. The five-number summary of year 2020 is as follows: min = -58.32, Q1 = -8.28, median = -3.87, Q3 = -1.10, max = 43.48
>>>>>>> d5445509dfbcdbde8fe01d26f22dfb942516fe03

### Part B

*Points: 2*

::: callout-note
2/2
:::

Calculate the mean and standard deviation of all countries' annual GDP growth rates in 2019. Do the same for 2020.

```{r}
library(tidyverse)
mean(gdp2019$value, na.rm=T)
sd(gdp2019$value, na.rm=T)
mean(gdp2020$value, na.rm=T)
sd(gdp2020$value, na.rm=T)
```

Mean of GDP growth in year 2019 is 2.89 with a s.d. of 3.45, while mean of GDP growth in year 2020 is -5.26 with a s.d. of 9.16

### Part C

*Points: 1*

::: callout-note
1/1
:::

Compare the mean and standard deviation of GDP growth rates in 2019 with those in 2020. What, generally, happened to each country's economy? Do not use the words or phrases "mean" or "standard deviation" in your answer.

In year 2020, the economy of all countries in general experienced a negative growth compared to that in year 2019. Also, the growth difference among individual countries tend to be larger in year 2020 than that in year 2019. In other words, the global economy went into recession with more fluctuations among countries.
