---
title: "Problem Set 4"
subtitle: "Due date: 9 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 30**

## Question 1

*Points: 5*

Suppose I am playing poker with a regular deck of 52 cards. I have three cards in my hand, one of which is a king. What is the probability that the next two cards I draw will both be kings?

```{r}
(3/49)*(2/48)
```

Answer: The probability of having the next two cards to both be kings is 0.26%.

## Question 2

*Points: 5*

A survey of 1,500 randomly selected individuals reveals that 46% of them have a college degree and that 31% of them have a gun in their house. If you were to assume that college graduation and gun ownership are independent of one another, what is the probability that an individual is both a college graduate and does not have a gun in their house?

```{r}
0.46*(1-0.31)
```

Answer: The probability of an individual to be both a college graduate and a non-gun owner is 31.74%.

## Question 3

*Points: 5*

I roll a fair six-sided dice and I get a number greater than three. What is the probability that the next two dice rolls will also generate numbers greater than three?

```{r}
(3/6)*(3/6)
```

Answer: The probability of having the next two rolls both with a number larger than three is 25%.

## Question 4

*Points: 10*

Assume that in the United States, 15% of people are smokers and, of smokers, 60% smoke at least one pack a day. Among those who smoke at least a pack a day, 60% will develop lung cancer at some point in their lives while, in the rest of the population (which includes smokers who smoke less than a pack a day and non-smokers), only 10% will. What is the probability that a person who develops lung cancer was a smoker who smoked at least a pack a day?

```{r}
# P(smoker) = 0.15, P(pack | smoker) = P(pack & smoker)/P(smoker) = 0.6
0.6*0.15
# Hence, P(pack & smoker) = 0.6*0.15 = 0.09
# ~P(pack & smoker) = 1-0.09 = 0.91, P(cancer & ~(pack & smoker)) = 0.091
# P (cancer|pack & smoker) = P (cancer & (pack & smoker))/P(pack & smoker) = 0.6
0.6*0.09
# Hence, P (cancer & (pack & smoker)) = 0.054
# Hence, P(cancer) = P(cancer & (pack & smoker)) + P(cancer & ~(pack & smoker)) = 0.091 + 0.054 = 0.145
# Finally, P((pack & smoker)|cancer) = P((pack & smoker) & cancer)/P(cancer)
0.054/0.145
```

Answer: The probability for a person who already has lung cancer to be a smoker that smokes at least a pack a day is 37.24%.

## Question 5

*Points: 5*

Assume that in the world, 60% of countries are democracies and 80% of countries have ratified the Rome Statute for the International Criminal Court. Assume further that 90% of democracies have ratified the Rome Accords. What is the probability that a country is either a democracy or has ratified the Rome Accords?

```{r}
# P(Dem) = 0.6, P(RS)=0.8, P(RS|Dem) = P(RS & Dem)/P(Dem) = 0.9
# Hence P (RS & Dem) = 0.9*0.6 = 0.54
# P (Dem OR RS) = P(Dem) + P(RS) - P(RS & Dem)
0.6+0.8-0.54
```

Answer: The probability of a country to be either a democracy or a Rome Accords ratified one is 86%.
