---
title: "Problem Set 5"
subtitle: "Due date: 23 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 25**

## Question 1

*Total points: 6*

Use the data in the table below to answer the following questions.

```{r}
#| echo: false

tibble::tibble(
  `Age group` = c("18-24", "25-30", "31 and up", "TOTAL"),
  `Non-voters` = c(70, 40, 220, 330),
  Voters = c(50, 50, 570, 670),
  Total = `Non-voters` + Voters
) |> 
  knitr::kable(caption = "Voting by Age in 2000")
```

### Part A

*Points: 2*

What is the probability of being 25-30 or a non-voter?

```{r}
(90/1000)+(330/1000)-(40/1000)
```

Answer: P(25-30 or non-voter) = P(25-30) + P(non-voter) - P (25-30 & non-voter) = 38%

### Part B

*Points: 4*

Assuming a normal distribution, report the 95% confidence intervals for the percentage of 18-to-24-year-olds who did not vote, and then the percentage of 25-to-30-year-olds who did not vote.

```{r}
70/120 # sample proportion p_hat
se_1 <- sqrt((70/120)*(1-70/120)/120) # s.e.
70/120 + 1.96*se_1
70/120 - 1.96*se_1
```

Answer_1: The 95% CIs for the percentage of 18-24 years old who did not vote is 0.495 to 0.671

```{r}
se_2 <- sqrt((40/90)*(1-40/90)/90) # s.e.
40/90 + 1.96*se_2
40/90 - 1.96*se_2
```

Answer_2: The 95% CIs for the percentage of 25-30 years old who did not vote is 0.342 to 0.547

## Question 2

*Total points:* *7*

Assume that the standard deviation for the population distribution of a state in which you want to conduct a poll is 200.

### Part A

*Points: 3*

Calculate the spread of the sampling distribution for each of the following sample sizes: 1, 4, 25, 100, 250, 1000, 5,000, and 10,000.

```{r}
# spread of the sampling distribution is the standard deviation of the sampling distrution, s.d.(samp_dist)= s.d.(pop_dist)/sqrt(n)
200/sqrt(1)
200/sqrt(4)
200/sqrt(25)
200/sqrt(100)
200/sqrt(250)
200/sqrt(1000)
200/sqrt(5000)
200/sqrt(10000)
```

Answer: The standard deviation of the sampling distribution is as follows: 200, 100, 40, 20, 12.65, 6.32, 2.83, and 2.

### Part B

*Points: 1*

Describe specifically how the variability of the sampling distribution changes as the sample size varies. Considering the expense of running a poll, which sample size do you think is most optimal if conducting the poll?

Answer: As the sample size expands, the variability of the sampling distribution goes down. For practical purposes (accuracy) of conducting a poll, 5,000 seems the best pick; while if there are further budgetary constraints, 1,000 may be the second best choice as the standard error is larger.

### Part C

*Points: 3*

Display your results graphically (using R) with the sample size on the x-axis and the standard error (of the sampling distribution) on the y-axis.

```{r}
library(ggplot2)

sample <- c(1, 4, 25, 100, 250, 1000, 5000, 10000)
sd <- 200
se <- sd / sqrt(sample)
data <- data.frame(x = sample, y = se)

ggplot(data, aes(x = x, y = y)) +
  geom_density(stat="identity") +
  labs(title = "Sample Size and Standard Error",
       x = "Sample size", y = "Standard error") +
  theme_minimal()
```

## Question 3

*Points: 4*

Suppose you conduct a survey (to generate a sample mean of interest) and find that it has a margin of error of 4.5 with a sample size of 900 using a 95% confidence interval. What would the margin of error be for a 90% confidence interval?

```{r}
1.645*(sqrt(900*(4.5^2)/(1.96^2))/sqrt(900))
```

Answer: The margin of error for a 90% CI is 3.78

## Question 4

*Points: 4*

Assume that, in State A, the mean income in the population is \$20,000 with a standard deviation of \$2,000. If you took an SRS of 900 individuals from that population, what is the probability that you would get a sample mean income of \$20,200 or greater? What would be the probability if the sample size was only 25?

::: callout-note
Assume a normal distribution for both questions.
:::

```{r}
se_41 <- 2000/sqrt(900)
se_42 <- 2000/sqrt(25)
z_41 <- (20200-20000)/se_41
z_42 <- (20200-20000)/se_42
# For a sample size of 900, P(X>20200) = P(Z>3) = 1-P(Z<=3) = 1-.9987 = 0.0013
# For a sample size of 25, P(X>20200) = P(Z>0.5) = 1-P(Z<=0.5) = 1-.6915 = 0.3085
# I got the normal distribution table from
# https://www.r-bloggers.com/2013/10/generating-your-own-normal-distribution-table/
```

Answer: For a sample size of 900, the probability of getting a sample mean larger than \$20,200 is 0.13%, while for a sample size of 25, the probability of getting a sample mean larger than \$20,200 is 30.85% (assuming normal distribution for both).

## Question 5

*Points: 4*

Assume that a coin is fair. If I flip a coin 500 times, what is a 95% confidence interval for the range of the count of heads that I will get? What if I flip the coin 5,000 times? What about 50,000 times?

```{r}
(0.5 + 1.96*sqrt((0.5)*(1-0.5)/500))*500
(0.5 - 1.96*sqrt((0.5)*(1-0.5)/500))*500
```

Answer_1: The 95% CI with 500 flips for heads is from 228 to 272 counts.

```{r}
(0.5 + 1.96*sqrt((0.5)*(1-0.5)/5000))*5000
(0.5 - 1.96*sqrt((0.5)*(1-0.5)/5000))*5000
```

Answer_2: The 95% CI with 5,000 flips for heads is from 2,431 to 2,569 counts.

```{r}
(0.5 + 1.96*sqrt((0.5)*(1-0.5)/50000))*50000
(0.5 - 1.96*sqrt((0.5)*(1-0.5)/50000))*50000
```

Answer_3: The 95% CI with 50,000 flips for heads is from 24,781 to 25,219 counts.
